Nitric oxide feedback to ciliary photoreceptor cells gates a UV avoidance circuit

### 2023-06-24 Manuscript/Manuscript.Rmd
MS: add math model methods, upload pics, general edits

Nitric oxide (NO) produced by nitric-oxide synthase (NOS) is a key regulator of animal physiology.
Here we uncover a function for NO in the integration of UV exposure and the gating of a UV-avoidance circuit.
We studied UV/violet avoidance mediated by brain ciliary photoreceptors (cPRCs) in larvae of the annelid *Platynereis dumerilii*. In the larva, NOS is expressed in interneurons (INNOS) postsynaptic to cPRCs.
UV stimulation of cPRCs triggers INNOS activation and NO production.
NO signals retrogradely to cPRCs to induce their sustained post-stimulus activation through an unconventional guanylyl cyclase.
This late activation inhibits serotonergic ciliomotor neurons to induce downward swimming.
In *NOS* mutants, retrograde signalling, circuit output and UV avoidance are defective.
By mathematical modelling, we recapitulate phototransduction and circuit dynamics in wild-type and mutant larvae.
Our results reveal how NO-mediated retrograde signalling gates a synaptic circuit and induces short-term memory of UV exposure to orchestrate light-avoidance behaviour.

@@

Here we uncovered an essential function for NO signalling in larval UV/violet-light avoidance behaviour.
In *Platynereis* larvae, UV/violet avoidance is mediated by brain ciliary photoreceptor cells (cPRCs) and is characterised by downward swimming [@Veraszto2018].
The cPRCs express a ciliary-type opsin, c-opsin1 [@Arendt2004] that forms a UV-absorbing bistable photopigment with an absorption maximum around 384 nm [@tsukamoto2017; @Veraszto2018; @Veedin2021].
Upon UV/violet exposure, the cPRC show a characteristic biphasic calcium response that is *c-opsin1* dependent.
UV/violet avoidance is also *c-opsin1*-dependent and is defective in *c-opsin1* mutants [@Veraszto2018].
Here we show that *Platynereis* *NOS* is expressed in interneurons of the cPRC circuit and is required for UV/violet-avoidance.
By combining calcium imaging across the fully-mapped cPRC circuit [@Veraszto2018] with genetic perturbations and mathematical modelling, we describe how NO tunes circuit dynamics through non-synaptic retrograde signalling to cPRCs.
This delayed neuroendocrine feedback integrates UV/violet exposure to induce a short-term memory manifested in altered circuit activity and an aversive behavioural response.

@@

We identified a single *nitric oxide synthase* (*NOS*) gene in the *Platynereis dumerilii* genome and transcriptome data.
Phylogenetic analysis of NOS proteins indicate that *Platynereis* NOS belongs to an orthology group of bilaterian NOS sequences (Figure 1---figure supplement 1).
To characterise the expression pattern of *NOS* we used in situ hybridization chain reaction (HCR) and transient transgenesis.
In two- and three-day-old larvae, we detected *NOS* expression in four cells (two of them weakly expressing) in the apical organ region (Figure 1D and Figure 1---figure supplement 2).
*NOS* was also expressed in the region of the visual eyes (adult eyes) and the pigmented eyespots (Figure 1---figure supplement 2).
The four apical organ cells, but not the eyes, were also labelled with a *NOS* reporter construct driving palmitoylated-tdTomato (Figure 1E).
This reporter also revealed the axonal projections of these central NOS-expressing neurons.
The position and morphology of the four *NOS+* cells allowed us to identify the same four cells as four interneurons (INNOS) in our three-day-old whole-body *Platynereis* volume EM data [@williams2017; @veraszto2020] (Figure 1B,C).
In the synaptic connectome, the INNOS cells are postsynaptic to the UV-sensory cPRCs and presynaptic to the INRGW interneurons, which are also cPRC targets (Figure 1C,F).
The projections of INNOS cells are segregated into input (dendritic) and output (axonal) compartments, with cPRC inputs occurring on the dendritic part in the dense neurosecretory plexus.
INNOS output synapses form in the more ventral projection region (Figure 1---figure supplement 3).
INRGW neurons synapse on the head serotonergic ciliomotor neurons (Ser-h1), which synapse on the prototroch ciliary band and the cholinergic MC ciliomotor neurons (Figure 1C,F) [@veraszto2017].

```{r echo=FALSE,fig.cap="**Figure 1. Identification of NOS-expressing interneurons (INNOS) within the cPRC circuit.** (**A**) Scanning electron microscopy image of a three-day-old *Platynereis* larva. (**B**, **C**) Volume rendering of the neuron types (cPRC, INNOS, INRGW, Ser-h1 and MC) in the cPRC circuit reconstructed from a whole-body transmission electron microscopy volume of a three-day-old larva. Neurite skeletons are shown with cell-body positions represented by spheres. Projections of all neurons in the body are shown in grey to highlight the neuropils. The outline of the yolk is also indicated in grey. In B, nuclei positions of the prototroch head ciliary band are shown as grey spheres. (**D**) Expression of the *NOS* gene detected by in situ HCR (magenta) in a two-day-old larva (anterior view). Antibody staining for acetylated α-tubulin (acTub: green) highlights cPRC cilia and the neuropil. (**E**) Expression of a *NOS* reporter (NOSp::palmi-3xHA: magenta) labelled with an anti-HA antibody in a two-day-old larva (anterior view). Antibody staining for acetylated α-tubulin (acTub: green) highlights cPRC cilia and the neuropil. (**F**) Synaptic wiring diagram of the cPRC circuit. Hexagons represent cell groups, with the number of cells per group shown in square brackets. Arrows represent the summed number of synaptic contacts between cell groups. Arrow thickness is proportional to the number of synapses."}

@@

**Two unconventional guanylyl cyclases are expressed in the cPRCs**

We aimed next to identify the NO receptor in the cPRCs.
NO generally acts via soluble guanylate cyclases (sGC), belonging to the guanylate cyclase family with a CYC domain (PFAM domain: PF00211).
NO binding to the heme group of sGC leads to increased cyclic guanosine monophosphate (cGMP) production.
Analysis of sGCs in *Platynereis* indicated that these genes are not expressed in any of the cells of the cPRC circuit [@veraszto2017].
Recently, Moroz and coworkers reported an atypical but widely conserved family of guanylyl cyclases with a NIT (nitrite/nitrate sensing) domain (PF08376) (NIT-GC) as potential mediators of NO signalling [@Moroz2020].
To identify NIT-GCs in *Platynereis*, we searched transcriptome resources and retrieved 15 potential NIT-GC homologs (Figure 4---figure supplement 1 and 2).
To analyse the relationship of these sequences to metazoan NIT-GCs, we retrieved protein sequences with a CYC domain from the transcriptome and genome databases of 45 metazoan and 2 choanoflagellate species.
We carried out cluster analysis and did phylogenetic reconstruction on a group of membrane-bound guanylyl cyclases with sGCs as an outgroup.
In agreement with Moroz et al. [@Moroz2020], we found a group of GCs with NIT domains with representatives in placozoans, cnidarians, some ecdysozoans, echinoderms, and lophotrochozoans.
The 15 *Platynereis* sequences belonged to several deeply diverged clades in the phylogenetic tree (Figure 4---figure supplement 1 and 2).

@@

To further characterise these atypical guanylyl cyclases, we focused on NIT-GC1 and carried out in vitro experiments.
In bacteria, NIT domains are thought to regulate cellular functions in response to intra- or extracellular nitrate and nitrite.
NIT-GC1 has a NIT domain and a highly conserved cyclase domain that is expected to catalyse cGMP synthesis (Figure 4G).
The NIT domain may render NIT-GC1 dependent on NO signals.
To test this, we co-expressed the cGMP indicator Green cGull [@Matsuda2017] and NIT-GC1 in cultured COS-7 (monkey kidney) cells, a cell line with minimal endogenous sGC activity [@Matsuda2017].
For balanced expression, we used a single plasmid with the two open-reading frames separated by the 2A self-cleaving peptide (Figure 4H).
Application of the NO donor SNAP lead to increased Green cGull fluorescence, an effect we did not observe when cells were exposed to DMSO or when Green cGull was expressed alone (Figure 4I-K).
To test whether this effect is dependent on the NIT domain, we also tested a deletion construct of NIT-GC1 lacking the NIT domain (Figure 4G).
Cells expressing this construct and Green cGull did not show an increased fluorescence of the cGMP reporter when exposed to SNAP (Figure 4L).
These results indicate that NIT-GC1 is able to catalyse cGMP production in an NO-dependent manner and this function requires the NIT domain.
These results establish NIT-GC1 as a biochemical sensor of NO or its derivatives.

@@

To investigate how NO and NIT-GC signalling influence the dynamics of the cPRC circuit, we imaged calcium signals from postsynaptic neurons in wild type, mutant and morphant larvae.
We were able to image the activity of all neurons in the cPRC circuit (INNOS, INRGW, Ser-h1 and MC).
The MC cell was identified based on its position and intrinsic activity [@veraszto2017].
To unambiguously identify all other cells from which we recorded calcium signals, we developed an on-slide immunostaining method (Figure 5---figure supplement 1A).
We used the cell-specific antibody markers against RYamide (INNOS) (Figure 5---figure supplement 1B-D), RGWamide (INRGW) and serotonin (Ser-h1) [@Conzelmann2011] to immunostain agar-embedded larvae following calcium imaging.
Based on the position of the nuclei, we could correlate live and fixed samples at a single-cell precision (Figure 5A,B).
Due to the stereotypy of the larvae, we could also identify neurons based on their position and calcium activity in activity-correlation maps (Figure 5C).

We first quantified the responses of the INNOS and INRGW interneurons during 405 nm stimulation of the cPRCs.
In both wild type and *NOS*-mutant larvae, INNOS cells showed an increase in calcium during stimulation (Figure 5D).
In contrast, the INNOS response was flat or slightly negative in NIT-GC2 morphant larvae (Figure 5E) revealing an essential role for NIT-GC2-mediated cPRC suppression in INNOS activation.
INRGW cells were initially inhibited during cPRC stimulation, followed by a delayed activation paralleling the second activation phase of cPRCs.
This late INRGW response was lacking in *NOS*-mutants (Figure 5F).
In NIT-GC2 morphants, INRGW cells showed a transient increase in calcium that decayed after light off and a delayed activation was not present (Figure 5G).

Next, we imaged calcium signals from Ser-h1 and MC neurons in wild type and *NOS* mutant larvae.
Ser-h1 cells showed an activation profile that correlated with cPRC activity, including a reduction in calcium during stimulation followed by rebound, a response that was defective in *NOS* mutants (Figure 5I).
MC cells showed sustained activation, including a late-phase that was lacking in *NOS* mutants (Figure 5I).
These data suggest that during 405 nm stimulation the Ser-h1 cells are inhibited and MC cells are activated, and this regulation is NO-dependent.
This pattern is expected to inhibit ciliary activity in the prototroch but not in the other ciliary bands, triggering NO-dependent downward swimming.

```{r echo=FALSE,fig.cap="**Figure 5. NOS- and NIT-GC2-dependent dynamics of the cPRC circuit.** (**A**, **B**) GCaMP6s imaging from cPRCs and INNOS cells (left panels) followed by on-slide immunostaining for (A) RYamide to label INNOS and (B) RGWamide+serotonin to label INRGW and Ser-h1 (red). Nuclei are stained with DAPI (cyan). Asterisks indicate cPRC nuclei. Numbers mark the same cells in the GCaMP and immunostaining images matched by position. (**C**) Correlation map of neuronal activity of the cPRCs, INNOS, INRGW, Ser-h1 and MC neurons. (**D**) GCaMP6s fluorescence in INNOS cells in wild type (WT) and NOSΔ11/Δ11 mutant larvae during 405 nm stimulation of the cPRC cilia. (**E**) GCaMP6s fluorescence in INNOS cells in NIT-GC2 morphant larvae during 405 nm stimulation. (**F**) GCaMP6s fluorescence in INRGW cells in wild type and NOSΔ23/Δ23 mutant larvae during 405 nm stimulation. (**G**) GCaMP6s fluorescence in INRGW cells in NIT-GC2 morphant larvae during 405 nm stimulation. (**H**, **I**) GCaMP6s fluorescence in (H) Ser-h1 cells and (I) the MC cell in wild type and NOSΔ11/Δ11 mutant larvae during 405 nm stimulation."}

@@

To further analyse the dynamics of responses to UV light and formally describe cPRC phototransduction, synaptic connections and NO retrograde signalling, we developed a mixed cellular-circuit-level mathematical model.
We used our Ca2+ imagining data of cPRC, INNOS and INRGW cells collected in wild type, *NOS* knockout, and *NIT-GC2* morphant larvae to formulate assumptions that are the basis of the proposed model.

We model a direct c-opsin1-dependent response to UV leading to the initial rise in Ca2+ levels.
The mechanism of this signal is not known but may include Gα and Gβ signalling, CNGAα channel activation or other pathways.
We further assume that in cPRC cells, Ca2+ levels increase with cGMP.
We model a decrease in cGMP and Ca2+ levels due to a NIT-GC2-dependent pathway (Figure 6A).
The effect of NO in cPRC cells is captured by another term that describes a NIT-GC1 and NO-dependent increase in cGMP.
Based on the synaptic connectome, we infer a feedforward coupling between cPRC cells and INNOS and INRGW cells.
The proposed synaptic coupling is UV-dependent, decays linearly and is suppressed in NIT-GC2 morphants (Figure 6A).

In the INNOS cells, we assume that excitatory synaptic input from cPRC (or rebound from tonic inhibition) leads to a rise in Ca2+ leading to NO production via NOS.
This input is NIT-GC2-dependent.
In the INRGW cells, we model a decrease in Ca2+ based on an inhibitory UV-dependent synaptic signal from the cPRCs.
We further assume the existence of direct excitatory coupling between cPRC Ca2+ levels and INRGW Ca2+ to account for the late effects of cPRC on INRGW.
The UV-dependent and Ca2+-dependent signals may be mediated by different transmitters released by cPRCs during distinct phases of their activation cycle.
In addition, we assume a direct inhibitory coupling between INNOS Ca2+ and INRGW Ca2+.
Finally, in all variables, we assumed a linear decay and constant production to set a steady state (Figure 6A).
Since the aim of the model is to capture the normalised fluorescence data, the model is nondimensionalised.
UV stimulation is modelled as a square pulse.
To find model parameters producing an output fitting the experimental data we employed a global optimisation method known as a genetic algorithm (GA).

Our model only includes a minimal set of parameters and assumptions of interactions that are required to describe the dynamics of cPRC, INNOS and INRGW in wild type and loss-of-function conditions (Figure 6A, B).
The model highlights that cPRC phototransduction employs distinct pathways that operate on different time scales and differentially influence cPRC Ca2+ levels.
The coupling between cPRCs and interneurons also requires different signalling mechanisms, either through distinct neurotransmitters or receptors expressed in the different cells.

To identify possible molecular pathways, we analysed single-cell transcriptome data for cPRC, INNOS and INRGW identified by spatial mapping [@Achim2015] and unique marker genes [@williams2017].
In cPRCs, we found transporters and synthetic enzymes indicating cholinergic, glutaminergic, glycinergic, GABAergic and adrenergic neurotransmission (Figure 6C).
For each neurotransmitter, we also found receptor-encoding genes expressed in INNOS and INRGW (Figure 6C).
The two types of interneurons often expressed different subunits or types of these receptors, indicating differential signalling (Figure 6C).

@@

@@

NO is a free radical with a millisecond-to-second half life and thus a limited signalling range.
In neuronal signalling, NOS is often localized to neurites [@Kuntz2017] at close proximity to sGC at synapses [@burette2002; @garthwaite2015a].
In the cPRC circuit, NOS is localised to the dendritic compartment of INNOS cells where also cPRC postsynaptic sites occur.
NIT-GC1, the target of NO signalling is localised to cPRC projections.
NO-mediated retrograde signalling thus likely occurs in the neurosecretory plexus where NOS- and NIT-GC1-containing projections are in close proximity and where we detected NO production following UV stimulation.
In contrast, INNOS to INRGW synapses occur outside the neurosecretory plexus in the ventral brain neuropil.

Compartmentalised signalling also occurs in peptidergic modulatory systems and can enable selective network activity during specific behaviours.
In the UV-avoidance circuit of *Drosophila* larvae, a peptidergic hub neuron Dp7 links UV-sensory neurons (v'td2) and motor circuits.
Dp7 expresses an an insulin-like peptide Ipl7 that is required for acute and sustained UV avoidance.
Ilp7 signalling occurs at a functionally and morphologically distinct dendritic compartment of Dp7, segregated from other sensory-motor pathways involving Dp7 [@Imambocus2022].

@@

Based on our data herein and previous work we can now propose a more detailed model of cPRC phototransduction and neurotransmission.
The cPRCs have high basal Ca2+ and respond to UV/violet light dependent on c-opsin1 [@Veraszto2018].
c-opsin1 forms a bistable photopigment and signals through Gi/oα and Gβγ [@tsukamoto2017; @Veedin2021; @tsukamoto2023].
In heterologous systems, the Gβγ subunits released following c-opsin1 activation open GIRK channels inducing K+ efflux (hyperpolarisation) [@tsukamoto2017; @tsukamoto2023] and close voltage-gated Ca2+ channels, thereby reducing intracellular calcium levels [@tsukamoto2023].
A GIRK channel is also expressed in the cPRCs (Figure 6A).
The pathway for the first rapid cPRC activation phase following UV/violet stimulus is not known but may involve the activation of a CNGα channel expressed in the cPRCs [@Tosches2014].
For the second inhibitory phase of phototransduction, we identified a key requirement for ciliary-localised NIT-GC2.
The mechanisms may involve c-opsin1-dependent inhibition of tonic NIT-GC2 activity and the reduction of ciliary cGMP.

>>>>

Nitric oxide (NO) produced by nitric-oxide synthase (NOS) is a key regulator of animal physiology.
Here we uncover a function for NO in the integration of UV exposure and the gating of a UV-avoidance circuit.
We studied UV/violet avoidance mediated by brain ciliary photoreceptors (cPRCs) in larvae of the annelid *Platynereis dumerilii*.
In the larva, NOS is expressed in interneurons (INNOS) postsynaptic to cPRCs.
UV stimulation of cPRCs triggers INNOS activation and NO production.
NO signals retrogradely to cPRCs to induce their sustained post-stimulus activation through an unconventional guanylate cyclase.
This late activation inhibits serotonergic ciliomotor neurons to induce downward swimming.
In *NOS* mutants, retrograde signalling, circuit output and UV avoidance are defective.
By mathematical modelling, we recapitulate phototransduction and circuit dynamics in wild-type and mutant larvae.
Our results reveal how NO-mediated retrograde signalling gates a synaptic circuit and induces short-term memory of UV exposure to orchestrate light-avoidance behaviour.

@@

We uncovered an essential function for NO signalling in larval UV/violet-light avoidance behaviour.
In *Platynereis* larvae, UV/violet avoidance is mediated by brain ciliary photoreceptor cells (cPRCs) and is characterised by downward swimming [@Veraszto2018].
The cPRCs express a ciliary-type opsin, c-opsin1 [@Arendt2004] that forms a UV-absorbing bistable photopigment with an absorption maximum around 384 nm [@tsukamoto2017; @Veraszto2018; @Veedin2021].
Upon UV/violet exposure, the cPRCs show a characteristic biphasic calcium response that is c-opsin1 dependent.
UV/violet avoidance is also *c-opsin1*-dependent and is defective in *c-opsin1* mutants [@Veraszto2018].
Here we show that *Platynereis* *NOS* is expressed in interneurons of the cPRC circuit and is required for UV/violet-avoidance.
By combining calcium imaging across the fully-mapped cPRC circuit [@Veraszto2018] with genetic perturbations and mathematical modelling, we describe how NO tunes circuit dynamics through non-synaptic retrograde signalling to cPRCs.
This delayed neuroendocrine feedback integrates UV/violet exposure to induce a short-term memory manifested in altered circuit activity and an aversive behavioural response.

@@

We identified a single *nitric oxide synthase* (*NOS*) gene in the *Platynereis dumerilii* genome and transcriptome data.
Phylogenetic analysis of NOS proteins indicate that *Platynereis* NOS belongs to an orthology group of bilaterian NOS sequences (Figure 1---figure supplement 1).
To characterise the expression pattern of *NOS* we used in situ hybridization chain reaction (HCR) and transient transgenesis.
In two- and three-day-old larvae, we detected *NOS* expression in four cells (two of them weakly expressing) in the apical organ region (Figure 1D and Figure 1---figure supplement 2).
*NOS* was also expressed in the region of the visual eyes (adult eyes) and the pigmented eyespots (Figure 1---figure supplement 2).
The four apical organ cells, but not the eyes, were also labelled with a *NOS* reporter construct driving palmitoylated-tdTomato (Figure 1E).
This reporter also revealed the axonal projections of these central NOS-expressing neurons.
The position and morphology of the four *NOS+* cells allowed us to identify the same four cells as four interneurons (INNOS) in our three-day-old whole-body *Platynereis* volume EM data [@williams2017; @veraszto2020] (Figure 1B,C).
In the synaptic connectome, the INNOS cells are postsynaptic to the UV-sensory cPRCs and presynaptic to the INRGWa interneurons, which are also cPRC targets (Figure 1C, F). 
The projections of INNOS cells are segregated into input (dendritic) and output (axonal) compartments, with cPRC inputs occurring on the dendritic part in the dense neurosecretory plexus.
INNOS output synapses form in the more ventral projection region (Figure 1---figure supplement 3).
INRGWa neurons synapse on the head serotonergic ciliomotor neurons (Ser-h1), which synapse on the cholinergic MC ciliomotor neuron and cells of the prototroch ciliary band (Figure 1C, F) [@veraszto2017].

```{r echo=FALSE,fig.cap="**Figure 1. Identification of NOS-expressing interneurons (INNOS) within the cPRC circuit.** **(A)** Scanning electron microscopy image of a three-day-old *Platynereis* larva. **(B, C)** Volume rendering of the neuron types (cPRC, INNOS, INRGWa, Ser-h1 and MC) in the cPRC circuit reconstructed from a whole-body transmission electron microscopy volume of a three-day-old larva. Neurite skeletons are shown with cell-body positions represented by spheres. Projections of all neurons in the body are shown in grey to highlight the neuropils. The outline of the yolk is also indicated in grey. In B, nuclei positions of the prototroch head ciliary band are shown as grey spheres. **(D)** Expression of the *NOS* gene detected by in situ HCR (magenta) in a two-day-old larva (anterior view). Antibody staining for acetylated α-tubulin (acTub: green) highlights cPRC cilia and the neuropil. **(E)** Expression of a *NOS* reporter (NOSp::palmi-3xHA: magenta) labelled with an anti-HA antibody in a two-day-old larva (anterior view). Antibody staining for acetylated α-tubulin (acTub: green) highlights cPRC cilia and the neuropil. **(F)** Synaptic wiring diagram of the cPRC circuit. Hexagons represent cell groups, with the number of cells per group shown in square brackets. Arrows represent the summed number of synaptic contacts between cell groups. Arrow thickness is proportional to the number of synapses."}

@@

**Two unconventional guanylate cyclases are expressed in the cPRCs**

We aimed next to identify the NO receptor in the cPRCs.
NO generally acts via soluble guanylate cyclases (sGC), belonging to the guanylate cyclase family with a CYC domain (PFAM domain: PF00211).
NO binding to the heme group of sGC leads to increased cyclic guanosine monophosphate (cGMP) production.
Analysis of sGCs in *Platynereis* indicated that these genes are not expressed in any of the cells of the cPRC circuit (not shown and [@veraszto2017]).
Recently, Moroz and coworkers reported an atypical but widely conserved family of guanylate cyclases with a NIT (nitrite/nitrate sensing) domain (PF08376) (NIT-GC) as potential mediators of NO signalling [@Moroz2020].
To identify NIT-GCs in *Platynereis*, we searched transcriptome resources and retrieved 15 potential NIT-GC homologs (Figure 4---figure supplement 1 and 2).
To analyse the relationship of these sequences to metazoan NIT-GCs, we retrieved protein sequences with a CYC domain from the transcriptome and genome databases of 45 metazoan and 2 choanoflagellate species.
We carried out cluster analysis and did phylogenetic reconstruction on a group of membrane-bound guanylate cyclases with sGCs as an outgroup.
In agreement with Moroz et al. [@Moroz2020], we found a group of GCs with NIT domains with representatives in placozoans, cnidarians, some ecdysozoans, echinoderms, and lophotrochozoans.
The 15 *Platynereis* sequences belonged to several deeply diverged clades in the phylogenetic tree (Figure 4---figure supplement 1 and 2).

@@

To further characterise these atypical guanylate cyclases, we focused on NIT-GC1 and carried out in vitro experiments.
In bacteria, NIT domains are thought to regulate cellular functions in response to intra- or extracellular nitrate and nitrite.
NIT-GC1 has a NIT domain and a highly conserved cyclase domain that is expected to catalyse cGMP synthesis (Figure 4G).
The NIT domain may render NIT-GC1 dependent on NO signals.
To test this, we co-expressed the cGMP indicator Green cGull [@Matsuda2017] and NIT-GC1 in cultured COS-7 (monkey kidney) cells, a cell line with minimal endogenous sGC activity [@Matsuda2017].
For balanced expression, we used a single plasmid with the two open-reading frames separated by the 2A self-cleaving peptide (Figure 4H).
Application of the NO donor SNAP lead to increased Green cGull fluorescence, an effect we did not observe when cells were exposed to DMSO or when Green cGull was expressed alone (Figure 4I-K).
To test whether this effect is dependent on the NIT domain, we also tested a deletion construct of NIT-GC1 lacking the NIT domain (Figure 4G).
Cells expressing this construct and Green cGull did not show an increased fluorescence of the cGMP reporter when exposed to SNAP (Figure 4L).
These results indicate that NIT-GC1 is able to catalyse cGMP production in an NO-dependent manner and this function requires the NIT domain.
These results establish NIT-GC1 as a biochemical sensor of NO or its derivatives.

@@

To investigate how NO and NIT-GC signalling influence the dynamics of the cPRC circuit, we imaged calcium signals from postsynaptic neurons in wild type, mutant and morphant larvae.
We were able to image the activity of all neurons in the cPRC circuit (INNOS, INRGWa, Ser-h1 and MC).
The MC cell was identified based on its position and intrinsic activity [@veraszto2017].
To unambiguously identify all other cells from which we recorded calcium signals, we developed an on-slide immunostaining method (Figure 5---figure supplement 1A).
We used the cell-specific antibody markers against RYamide (INNOS) (Figure 5---figure supplement 1B-D), RGWamide (INRGWa) and serotonin (Ser-h1) [@Conzelmann2011] to immunostain agar-embedded larvae following calcium imaging.
Based on the position of the nuclei, we could correlate live and fixed samples at a single-cell precision (Figure 5A, B).
Due to the stereotypy of the larvae, we could also identify neurons based on their position and calcium activity in activity-correlation maps (Figure 5C).

We first quantified the responses of the INNOS and INRGWa interneurons during 405 nm stimulation of the cPRCs.
In both wild type and *NOS*-mutant larvae, INNOS cells showed an increase in calcium during stimulation (Figure 5D).
In contrast, the INNOS response was flat or slightly negative in NIT-GC2 morphant larvae (Figure 5E) revealing an essential role for NIT-GC2-mediated cPRC suppression in INNOS activation.
INRGWa cells were initially inhibited during cPRC stimulation, followed by a delayed activation paralleling the second activation phase of cPRCs.
This late INRGWa response was lacking in *NOS*-mutants (Figure 5F).
In NIT-GC2 morphants, INRGWa cells showed a transient increase in calcium that decayed after light off and a delayed activation was not present (Figure 5G).

Next, we imaged calcium signals from Ser-h1 and MC neurons in wild type and *NOS* mutant larvae.
Ser-h1 cells showed an activation profile that correlated with cPRC activity, including a reduction in calcium during stimulation followed by rebound, a response that was defective in *NOS* mutants (Figure 5I).
MC cells showed sustained activation, including a late-phase that was lacking in *NOS* mutants (Figure 5I).
These data suggest that during 405 nm stimulation the Ser-h1 cells are inhibited and MC cells are activated, and this regulation is NO-dependent.
This pattern is expected to inhibit ciliary activity in the prototroch but not in the other ciliary bands (with no Ser-h1 or MC synapses), triggering NO-dependent downward swimming.

```{r echo=FALSE,fig.cap="**Figure 5. NOS- and NIT-GC2-dependent dynamics of the cPRC circuit.** **(A, B)** GCaMP6s imaging from cPRCs and INNOS cells (left panels) followed by on-slide immunostaining for (A) RYamide to label INNOS and (B) RGWamide+serotonin to label INRGWa and Ser-h1 (red). Nuclei are stained with DAPI (cyan). Asterisks indicate cPRC nuclei. Numbers mark the same cells in the GCaMP and immunostaining images matched by position. **(C)** Correlation map of neuronal activity of the cPRCs, INNOS, INRGWa, Ser-h1 and MC neurons. **(D)** GCaMP6s fluorescence in INNOS cells in wild type (WT) and NOSΔ11/Δ11 mutant larvae during 405 nm stimulation of the cPRC cilia. **(E)** GCaMP6s fluorescence in INNOS cells in NIT-GC2 morphant larvae during 405 nm stimulation. **(F)** GCaMP6s fluorescence in INRGWa cells in wild type and NOSΔ23/Δ23 mutant larvae during 405 nm stimulation. **(G)** GCaMP6s fluorescence in INRGWa cells in NIT-GC2 morphant larvae during 405 nm stimulation. **(H, I)** GCaMP6s fluorescence in (H) Ser-h1 cells and (I) the MC cell in wild type and NOSΔ11/Δ11 mutant larvae during 405 nm stimulation."}

@@

To further analyse the dynamics of responses to UV light and formally describe cPRC phototransduction, synaptic connections and NO retrograde signalling, we developed a mixed cellular-circuit-level mathematical model.
We used our Ca^2+^ imagining data of cPRC, INNOS and INRGWa cells collected in wild type, *NOS* knockout, and *NIT-GC2* morphant larvae to formulate assumptions that are the basis of the proposed model.

We model a direct c-opsin1-dependent response to UV leading to the initial rise in Ca^2+^ levels.
The mechanism of this signal is not known but may include Gα and Gβ signalling, CNGAα channel activation or other pathways.
We further assume that in cPRC cells, Ca^2+^ levels increase with cGMP.
We model a decrease in cGMP and Ca^2+^ levels due to a NIT-GC2-dependent pathway (Figure 6A).
The effect of NO in cPRC cells is captured by another term that describes a NIT-GC1 and NO-dependent increase in cGMP.
Based on the synaptic connectome, we infer a feedforward coupling between cPRC cells and INNOS and INRGWa cells.
The proposed synaptic coupling is UV-dependent, decays linearly and is suppressed in NIT-GC2 morphants (Figure 6A).

In the INNOS cells, we assume that excitatory synaptic input from cPRC (or rebound from tonic inhibition) leads to a rise in Ca^2+^ leading to NO production via NOS.
This input is NIT-GC2-dependent.
In the INRGWa cells, we model a decrease in Ca^2+^ based on an inhibitory UV-dependent synaptic signal from the cPRCs.
We further assume the existence of direct excitatory coupling between cPRC Ca^2+^ levels and INRGWa Ca^2+^ to account for the late effects of cPRC on INRGWa (Figure 6A).
The UV-dependent and Ca^2+^-dependent signals may be mediated by different transmitters released by cPRCs during distinct phases of their activation cycle.
In addition, we assume a direct inhibitory coupling between INNOS Ca^2+^ and INRGWa Ca^2+^.
Finally, in all variables, we assumed a linear decay and constant production to set a steady state (Figure 6A).
Since the aim of the model is to capture the normalised fluorescence data, the model is nondimensionalised.
UV stimulation is modelled as a square pulse.
To find model parameters producing an output fitting the experimental data we employed a global optimisation method known as a genetic algorithm (GA).

Through this optimisation procedure, we found parameters that gave a very good fit to our experimental data under all conditions.
We could fit to mean Ca^2+^ values but also individual calcium transients (Figure 6B, Figure 6--figure supplement 1).
Our model only includes a minimal set of parameters and assumptions of interactions that are required to describe the dynamics of cPRC, INNOS and INRGWa in wild type and loss-of-function conditions (Figure 6A, B).
The model highlights that cPRC phototransduction employs distinct pathways that operate on different time scales and differentially influence cPRC Ca^2+^ levels.
The coupling between cPRCs and interneurons also requires different signalling mechanisms, either through distinct neurotransmitters or receptors expressed in the different cells.

To identify possible molecular pathways, we analysed single-cell transcriptome data for cPRC, INNOS and INRGWa identified by spatial mapping [@Achim2015] and unique marker genes [@williams2017].
In cPRCs, we found transporters and synthetic enzymes indicating cholinergic, glutaminergic, glycinergic, GABAergic and adrenergic neurotransmission (Figure 6C).
For each neurotransmitter, we also found receptor-encoding genes expressed in INNOS and INRGWa (Figure 6C).
The two types of interneurons often expressed different subunits or types of these receptors, indicating differential signalling (Figure 6C).

@@

To generate new insights from the model, we asked how the duration and intensity of UV stimulation influences signalling output, measured as the magnitude (quantified as area under peak, AUP) of the NO-dependent Ca^2+^ signal in the cPRCs.
The model allowed us to systematically vary stimulus conditions that would not have been feasible in single-larva experiments.
We found that delayed activation occurs only for UV stimulation of sufficient amplitude and  duration (Figure 7).
This suggests an integratory function of NO signalling to tune circuit output to the strength of UV exposure.

```{r echo=FALSE,fig.cap="**Figure 7. Magnitude of cPRC activation as  a function of stimulus duration and intensity.** Median values of the $AUP$ are presented as a heatmap with dark blue indicating flat $C_P(t)$ trajectories with median $AUP$ close to 0 and red indicating that the $C_P(t)$ trajectories have a clearly defined delayed activation peak. "}

knitr::include_graphics(path = "figures/Fig7.svg")
```

@@

NO is a free radical with a millisecond-to-second half life and thus a limited signalling range.
In neuronal signalling, NOS is often localized to neurites [@Kuntz2017] at close proximity to sGC at synapses [@burette2002; @garthwaite2015a].
In the cPRC circuit, NOS is localised to the dendritic compartment of INNOS cells where also cPRC postsynaptic sites occur.
NIT-GC1, the target of NO signalling is localised to cPRC projections.
NO-mediated retrograde signalling thus likely occurs in the neurosecretory plexus where NOS- and NIT-GC1-containing projections are in close proximity and where we detected NO production following UV stimulation.
In contrast, INNOS to INRGWa synapses occur outside the neurosecretory plexus in the ventral brain neuropil.

Compartmentalised signalling also occurs in peptidergic modulatory systems and can enable selective network activity during specific behaviours.
In the UV-avoidance circuit of *Drosophila* larvae, a peptidergic hub neuron Dp7 links UV-sensory neurons (v'td2) and motor circuits.
Dp7 expresses an insulin-like peptide Ilp7 that is required for acute and sustained UV avoidance.
Ilp7 signalling occurs at a functionally and morphologically distinct dendritic compartment of Dp7, segregated from other sensory-motor pathways involving Dp7 [@Imambocus2022].

@@

Based on our data herein and previous work we can now propose a more detailed model of cPRC phototransduction and neurotransmission.
The cPRCs have high basal Ca^2+^ and respond to UV/violet light dependent on c-opsin1 [@Veraszto2018].
c-opsin1 forms a bistable photopigment and signals through Gi/oα and Gβγ [@tsukamoto2017; @Veedin2021; @tsukamoto2023].
In heterologous systems, the Gβγ subunits released following c-opsin1 activation open GIRK channels inducing K+ efflux (hyperpolarisation) [@tsukamoto2017; @tsukamoto2023] and close voltage-gated Ca^2+^ channels, thereby reducing intracellular calcium levels [@tsukamoto2023].
A GIRK channel is also expressed in the cPRCs (Figure 6A).
The pathway for the first rapid cPRC activation phase following UV/violet stimulus is not known but may involve the activation of a CNGα channel expressed in the cPRCs [@Tosches2014].
For the second inhibitory phase of phototransduction, we identified a key requirement for ciliary-localised NIT-GC2.
The mechanisms may involve c-opsin1-dependent inhibition of tonic NIT-GC2 activity and the reduction of ciliary cGMP.
